---
title: "Comparing outputs from lefser, lefse-conda, and lefse-galaxy using the same dataset (zeller14)"
output:
    html_document: 
        toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(lefser)
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)
library(devtools)
library(tibble)
library(tidyr)
library(magrittr)
library(readr)
library(VennDiagram)
library(purrr)
```

# Introduction

In this document, I compare the output from lefse through different
applications:

+ lefser (R package)
+ lefse-conda (command line)
+ lefse-galaxy (from the galaxy platfrom)

In all cases, I use the same dataset, zerrer14, which is included in the
zeller14 package.

# Run lefse independently with the three applications (R, conda, galaxy)

## Running lefse in R (lefser)

Prepare data with TSS normalization. Lefse uses this same normalization step.

```{r}
data('zeller14')
se <- zeller14[,zeller14$study_condition != 'adenoma']
rownames(se) <- sub('^.+([a-z]__.+$)', '\\1', rownames(se))

colData(se)[['study_condition']] <- factor(
    colData(se)[['study_condition']], levels = c('control', 'CRC')
)

abundances <- assay(se)
abundances_tss <- apply(abundances, 2, function(x) x / sum(x) * 1e6)
assays(se)[[2]] <- abundances_tss
names(assays(se))[2] <- "tss"
se
```

Perform the analysis with the `lefser` function:

```{r}
lefser_output <- lefser(
    se, kruskal.threshold = 0.05, wilcox.threshold = 0.05,
    groupCol = 'study_condition', blockCol = 'age_category', assay = 2L
) %>% 
    mutate(app_name = 'lefser') %>% 
    arrange(scores) %>% 
    rename(feature = Names, lefser_LDA = scores)

head(lefser_output)

```

## Running lefse-conda (command line) 

### lefse-conda installation and version

Note: I installed lefse following the instructions from [this site](https://github.com/biobakery/conda-biobakery)
after installing conda.

```{bash, eval=FALSE}
## Add channels
conda config --add channels defaults
conda config --add channels bioconda
conda config --add channels conda-forge
conda config --add channels biobakery

## Install lefse
conda create -n lefse
conda activate lefse
conda install -c biobakery lefse
```

Conda and lefse versions:

```{bash, eval=FALSE}
conda --version
#> conda 4.11.0

conda list | grep -e "lefse"
# packages in environment at /home/samuel/miniconda3/envs/lefse:
#> lefse                     1.1.2              pyhdfd78af_0    bioconda

```

### Run lefse-conda

1. Generate a tabular dataset (zeller14) compatible with lefse-conda and
lefse-galaxy using the `get_dataset.R` script.

2. Run the script `run_lefse.sh` (linux) with the following parameters:

```{bash, eval=FALSE}
## In general
## ./run_lefse.sh <path/to/conda/activate> <env_name> <path/to/Rscript>

## in my case (Samuel)
./run_lefse.sh ~/miniconda3/bin/activate lefse ~/bin/Rscript
```

Note: All script files, `get_dataset.R` and `run_lefse.sh`, and this rmarkdown
document must be in the same directory.

### Import output from lefse-conda into R

```{r}
col_names <- c(
    'feature', 'log_hi_class_avg', 'class', 'lefse_conda_LDA', 'pval'
)
lefse_conda <- readr::read_tsv(
    'zeller14.res', show_col_types = FALSE, col_names = FALSE 
) %>% 
    set_colnames(col_names) %>% 
    filter(!is.na(lefse_conda_LDA)) %>%
    mutate(
        lefse_conda_LDA = ifelse(
            class == 'control', -lefse_conda_LDA, lefse_conda_LDA
        ),
        app_name = 'lefse_conda'
    ) %>% 
    arrange(lefse_conda_LDA)
    
head(lefse_conda)
```

## Running lefse from galaxy

I used the `zeller14_nosub.txt` file (no subjects included) as input for
lefse from the galaxy platform of the Huttenhower lab at 
https://huttenhower.sph.harvard.edu/galaxy/

alpha were 0.05 for both KW and Wilcox, 2.0 for LDA. TSS normalization was
applied as well.

I'll just import the output:

```{r}
col_names <- c(
    'feature', 'log_hi_class_avg', 'class', 'lefse_galaxy_LDA', 'pval'
)
lefse_galaxy <- read_tsv(
    "lefse_gal_all_vs_all.res", col_names = FALSE, show_col_types = FALSE
) %>% 
    set_colnames(col_names) %>% 
    filter(!is.na(lefse_galaxy_LDA)) %>% 
    mutate(
        lefse_galaxy_LDA = ifelse(
            class == "control", -lefse_galaxy_LDA, lefse_galaxy_LDA
        ),
        app_name = 'lefse_galaxy'
    )

head(lefse_galaxy)
```

# Comparison of lefse-conda with lefser

## Number of features reported as significant

```{r}
combined_outputs <- bind_rows(lefse_conda, lefse_galaxy, lefser_output) %>% 
    mutate(LDA = coalesce(lefse_conda_LDA, lefse_galaxy_LDA, lefser_LDA))

combined_outputs %>% 
    count(app_name) %>% 
    ggplot(aes(app_name, n)) +
    geom_col() +
    geom_label(aes(label = n)) +
    ggtitle(
        'Number of significiant features identified by the different applications using lefse'
    )
```


## Overlap of features reported as significant

```{r}
set1 = lefse_conda$feature
set2 = lefse_galaxy$feature
set3 = lefser_output$feature

grid.newpage()
venn_object <- venn.diagram(
    x = list(set1, set2, set3),
    category.names = c("lefse-conda" , "lefse-galaxy" , "lefser"),
    filename = NULL
)
grid.draw(venn_object)

```

## LDA scores are similar

LDA scores of the 24 overlapping features are similar.

```{r}
joint_output <- purrr::reduce(
    .x = list(lefse_conda, lefse_galaxy, lefser_output),
    .f = ~ inner_join(.x, .y, by = 'feature')
) %>% 
    select(feature, lefse_conda_LDA, lefse_galaxy_LDA, lefser_LDA)
joint_output
```

### lefser LDA scores vs lefse-conda LDA scores

```{r}
joint_output %>% 
    ggplot(aes(lefse_conda_LDA, lefser_LDA)) + 
    geom_point(size = 3, shape = 1) +
    geom_hline(yintercept = 0, linetype = 'dashed') +
    geom_vline(xintercept = 0, linetype = 'dashed') +
    ggtitle(
    'Comparison of LDA scores of features reported as significant 
    by both lefse-conda and lefser'
    )
```


### lefser LDA scores vs lefse-galaxy LDA scores

```{r}
joint_output %>% 
    ggplot(aes(lefse_galaxy_LDA, lefser_LDA)) + 
    geom_point(size = 3, shape = 1) +
    geom_hline(yintercept = 0, linetype = 'dashed') +
    geom_vline(xintercept = 0, linetype = 'dashed') +
    ggtitle(
    'Comparison of LDA scores of features reported as significant 
    by both lefse-galaxy and lefser'
    )
```


# Session info

```{r}
session_info()
```

